\chapter{Greetings and Expressions}

When people meet, there are a number of greetings that are employed. One must ask about a person’s children and family.
\\
\gll úləlít̪ano\\
\\
	\trans	morning	(the word for morning used as a greeting in the morning)\\
\gll ŋót̪áu		\\
	\\
	\trans		hello, how are you?	(used as a general greeting)\\
\gll ɲuŋgətiə		\\
	\\
	\trans		I am fine	(WHAT verb is this?)\\
\gll ðaó ŋopea		\\
	\\
	\trans		it is good/well\\
\gll égaó məna		\\
	\\
	\trans		I am not right (in health)\\
\gll aˠaó mənːaŋ		\\
	\\
	\trans		 you are not right (in health)\\
\gll dáó ŋopea		\\
	\\
	\trans		be well! (sit well)		CHECK\\
\gll ani ɲágaŋəɾá?	\\
	\\
	\trans			how are you (all)?\\
\gll ɲágaŋəɾa		\\
	\\
	\trans		we are fine\\
\gll ani ɲeɾá ɲaŋəɾá?\\
	\\
	\trans		how are the children?\\
\gll ɲɜ́tu t̪iə ɲaŋəɾá	\\
	\\
	\trans		they are fine\\
\gll ani ɲaŋeɾá ʧom éɲəɲe\\
	\\
	\trans	are yours fine too?\\
\gll ʤuliə kadó t̪áu	\\
	\\
	\trans		Julia, how is she? \\
\gll kɜ́tu t̪iə		\\
	\\
	\trans		she is fine\\


\subsection{Interjections and exclamations}
\begin{tabular}[t]{ll}
nda 	&	no\\
la		&	no  (from Arabic)\\
ai		&	yes\\
\end{tabular}
\\
\gll ówːa kapːó lóŕná ? 		\\
\\
\trans		did the woman pick up the basket?\\
\gll ai, kapːó lóŕná 		\\
\\
\trans			yes, she picked up the basket\\
\gll la, ówːa kanːa áŋə́pa lóŕná kapːó lə́búŋwá\\
\\
\trans	no, the woman did not pick up the basket, she picked up the waterpot\\
\gll wɜj		\\
\\
\trans	woah! this is an exclamation of surpise, distaste\\
\gll wɜj, ekegəŋe	\\
\\
\trans	woah, the bitterness (what a bad thing!)	\hfill 6/26/2013\\


\subsubsection{árrá}\section{sec:ch21:vocative}
This is used to attract attention or to point something out, like ‘hey’ in English. The focus of attention is marked with a final -u or -w. 
\\
\gll árrá kúkə-w\\
\\
\trans	behold Kuku or hey, look at Kuku!			\hfill 6/26/2013\\
\\
\gll árrá	íɲɲé-w 	ɲ-eɾt̪-ó 			iɾiə	 	joaɲa	\\
	hey  	1\textsc{sg.pro-emph}	1\textsc{sg.sm}-have-\textsc{pfv}  	\textsc{cl}j.cow  	\textsc{cl}j.many\\
\trans Hey, as for me, I have so many cows!\\
\\
\gll pi	\\
\\
\trans exclamation of surprise\\
\gll ɜt:iə	\\
\\
\trans ‘is that so?’\\

